---
title: "French trains"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(scales)
```

```{r}
small_trains <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-26/small_trains.csv")

small_trains <- small_trains %>% 
  mutate(
    arrival_station = str_to_title(arrival_station),
    departure_station = str_to_title(departure_station)
  )

full_trains <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-26/full_trains.csv")

full_trains <- full_trains %>% 
  mutate(
    arrival_station = str_to_title(arrival_station),
    departure_station = str_to_title(departure_station),
    date = as.Date(paste(year, month, 1, sep="/"))
  )
```

```{r}
str(full_trains)
```

Each row is a monthly summary of train routes, the number of routes is pretty stable for 2015-2017, but more are added in 2018.

```{r}
small_trains %>% 
  count(year, month)
```

```{r}
full_trains %>% 
  count(year, month)
```

The small train dataset has 6 times as many rows as the full dataset. Based on the tidy-tuesday README this seems to be because delay causes are gathered for the small trains data. Can we prove this by picking a particular route, e.g. Paris Est to Metz in a particular month

```{r}
small_trains %>% 
  filter(departure_station == "Paris Est", arrival_station == "Metz", year == 2017, month == 9)
```

```{r}
full_trains %>% 
  filter(departure_station == "Paris Est", arrival_station == "Metz", year == 2017, month == 9) %>% 
  select(year, month, departure_station, arrival_station, contains("delay"))
```

# Which routes have the longest journeys?

```{r}
route_journey_times <- full_trains %>% 
  group_by(departure_station, arrival_station) %>% 
  summarise(
    journey_time_avg = mean(journey_time_avg, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  mutate(
    journey_time_avg_overall = mean(journey_time_avg),
    journey_time_diff = journey_time_avg - journey_time_avg_overall,
    route = paste(departure_station, arrival_station, sep = " to ")
  ) %>% 
  select(route, journey_time_avg, journey_time_diff)

longest_routes <- route_journey_times %>% 
  top_n(5, wt = journey_time_avg)
  
shortest_routes <- route_journey_times %>%
  top_n(-5, wt = journey_time_avg)

top_bottom <- bind_rows(longest_routes, shortest_routes)

top_bottom %>% 
  mutate(route = fct_reorder(route, journey_time_diff)) %>% 
  arrange(journey_time_diff) %>% 
  ggplot(aes(route, journey_time_avg, fill = journey_time_diff > 0)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11),
    panel.grid.major.y = element_blank()
  ) +
  labs(
    title = "Longest and shortest train journeys",
    subtitle = "Compared to average journey time",
    x = "",
    y = "Difference from average (minutes)"
  )
```


# Which station has the worst punctuality?

```{r}
fill_colours_manual = c("azure3", "deepskyblue4")
theme_set(
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank()
  )
)

top_delay_stations <- full_trains %>%
  group_by(departure_station) %>% 
  summarize(
    total_trips = sum(total_num_trips),
    delays = sum(num_late_at_departure)
  ) %>% 
  ungroup() %>% 
  mutate(
    overall_num_delays = sum(delays),
    percent_all_delays = delays / overall_num_delays,
    lyon = ifelse(departure_station == "Lyon Part Dieu", T, F)
  ) %>% 
  mutate(
    departure_station = fct_reorder(departure_station, percent_all_delays)
  ) %>% 
  top_n(10, percent_all_delays)

top_delay_stations %>% 
  mutate(departure_station = fct_reorder(departure_station, total_trips)) %>% 
  ggplot(aes(departure_station, total_trips, fill = lyon)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = comma_format()) +
  scale_fill_manual(values = fill_colours_manual) +
  labs(
    title = "Total number of trips",
    subtitle = "Lyon Part Dieu is the 3rd busiest station",
    x = "",
    y = "Total trips"
  )

ggsave("top_10_trips_plot.png")

top_delay_stations %>% 
  ggplot(aes(departure_station, percent_all_delays, fill = lyon)) +
  geom_col() +
  geom_text(
    aes(label = paste0(round(percent_all_delays * 100, 1),"%")), 
    fontface = "bold", 
    color = "white", 
    hjust=1.2
  ) +
  coord_flip() +
  scale_y_continuous(labels = percent_format()) + 
  scale_fill_manual(values = fill_colours_manual) +
  labs(
    title = "Proportion of all delays",
    subtitle = "Lyon Part Dieu accounts for the most delays",
    x = "",
    y = "Delayed (%)"
  )

ggsave("proportion_of_delayed_trips.png")
```


